---
title: "solutions"
author: "Dhairya Dalal"
date: "May 29, 2015"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r, echo=FALSE, include=FALSE}
######## [Load Libraries] #########
library(dplyr)
library(ggplot2)
library(knitr)
library(xtable)
library(ggvis)
library(statebins)

######## [Load Data] #########
# 1. Read school_data.csv using read.csv() and save results to data
sdata = read.csv(file = "data/school_data.csv", stringsAsFactors = FALSE, na.strings = "")

#2. Convert all column names to lower case
names(sdata) = tolower(names(sdata))

#3. Strip dollar signs from award_amount_2010-2011 columm. More tips on data cleaning here: http://earlh.com/blog/2009/06/29/cleaning-data-in-r-csv-files/
#a. Use gsub() and regeular expression to strip extraneous characters
sdata$award_amount_2010.2011 = gsub('[^a-zA-Z0-9.]', '', sdata$award_amount_2010.2011)
#b. Set class of column to numeric
sdata$award_amount_2010.2011 = as.numeric(sdata$award_amount_2010.2011)
```


## Counting things
How many schools applied for a grant? 
```{r}
# a. Let's use the length() function to count the numbers of schools 
#     in our dataset.
# b. In order get the column with schools names try using the dollar 
#    sign subselector (dataframe$column). For example the schools in 
#    our dataset would be be sdata$schoo_name
school_count = length(sdata$school_name)
```
Answer: `r school_count` schools applied for grants in 2010-2011. 

## Counting unique things pt 2
How many unique schools applied for a grant? 
```{r}
#a. To accomplish this we'll use the unique() function
unique_schools = unique(sdata$school_name)
lous = length(unique_schools)

# You can can also express this in one line using a nested function call
lous = length(unique(sdata$school_name))
```
Answer: `r lous` schools applied for grants in 2010-2011. 

## Simple summary Statistics - Sum
What was the total amount of grant dollars disbursed? 
```{r}
#a. To accomplish this we'll use sum function
total = sum(sdata$award_amount_2010.2011)
total

#b. NA, that doesn't seem right. Let's ignore the 
#  NA values in our total calculation using the 
#  argument na.rm = TRUE
total = sum(sdata$award_amount_2010.2011, na.rm = TRUE)

#c. Formatted total
formatted_total = format(total, big.mark = ",", small.mark = ",")
```
Answer: $`r formatted_total` was disbursed in 2010 - 2011. 

## Simple summary Statistics - Mean
What was average amount of grant dollars disbursed? 
```{r}
#a. To accomplish this we'll use the mean funtion. Don't
#   forget to use na.rm=TRUE. 
avg = mean(sdata$award_amount_2010.2011, na.rm = TRUE)

#b. Formatted avg
formatted_avg = format(avg, big.mark = ",", small.mark = ",", nsmall = 2)
```
Answer: On average, $`r formatted_avg` in grant dollars was disbursed. 

## Simple summary Statistics - Median
What was the median amount of grant dollars disbursed?
```{r}
#a. To accomplish this we'll use the median funtion. Don't
#   forget to use na.rm=TRUE. 
med = median(sdata$award_amount_2010.2011, na.rm = TRUE)

#b. Formatted avg
formatted_med = format(med, big.mark = ",", small.mark = ",", nsmall = 2)
```
Answer: The median is: $`r formatted_med`. 

## Simple summary Statistics - Minimum
What was the smallest amount of grant dollars disbursed?
```{r}
#a. To accomplish this we'll use the min() funtion. Don't
#   forget to use na.rm=TRUE. 
min_ = min(sdata$award_amount_2010.2011, na.rm = TRUE)

#b. Formatted avg
formatted_min = format(min_, big.mark = ",", small.mark = ",", nsmall = 2)
```
Answer: The smallest amount granted was $`r formatted_min`. 

## Simple summary Statistics - Max
What was the median amount of grant dollars disbursed?
```{r}
#a. To accomplish this we'll use the median funtion. Don't
#   forget to use na.rm=TRUE. 
max_ = max(sdata$award_amount_2010.2011, na.rm = TRUE)

#b. Formatted avg
formatted_max = format(max_, big.mark = ",", small.mark = ",", nsmall = 2)
```
Answer: The largest amount granted was $`r formatted_max`.

## Intro to dplyr: Filtering and selecting {.smaller}
Which schools recieved more than $900,000?
```{r}
# We'll use the filter function in our dplyr chain
nkschools = sdata %>% 
            filter(award_amount_2010.2011 > 900000) %>%
            select(state, school_name, award_amount_2010.2011) %>%
            arrange(desc(award_amount_2010.2011)) %>% head(5)

kable(nkschools)
```

## Intro to dplyr: Aggregation and Summarization {.smaller}
What are the top 3 states which recieved the most funding overall?
```{r xtable, results="asis"}
# Dplyr logic
# group data by states > calculate total grant dollars per state 
funding_state = sdata %>% 
                group_by(state) %>% 
                summarise(total = sum(award_amount_2010.2011, na.rm = TRUE))

# Top 5 States in overall funding
ordered_states = funding_state %>% arrange(desc(total))
ordered_states$total = format(ordered_states$total, big.mark = ",", small.mark = ",", nsmall = 2)
top_3 = head(ordered_states, 3)
kable(top_3)
```

## Intro to dplyr: Aggregation and Summarization {.smaller}
What are the bottom 5 states which recieved the least funding overall?
```{r}
# Bottom 10 states in overall funding
bottom_5 = tail(ordered_states, 5)
kable(bottom_5)
```

## Data Visualization - Scatter plot
```{r}
# Let's try to quickly plot our dataset using a scatter plot
qplot(state, award_amount_2010.2011, data = sdata, color = state)
```

## Interlude - Joins
```{r warning=FALSE}
# Whoa, that was unwieldy. There are so many states. Let's try map each state 
# to a region and replot each grant by region.
#a. First let's read ing the geocodes dataset
geocodes = read.csv(file="data/geocodes.csv")
head(geocodes,3) # Take quick peek at the new table

#b. Next let's join sdata and geocodes by state and code
joined = left_join(sdata, geocodes, by = c("state" = "Code"))
head(joined, 1)
```

## Data Visualization - Scatter plot
```{r echo = FALSE, warning=FALSE, fig.width=10}
# Let's try to replot our dataset using region instead
library(ggthemes)
options(scipen=999)
qplot(state, award_amount_2010.2011, data = joined, color = Region) +
     theme(axis.text.x = element_text(angle = 70, hjust = 1, vjust = 1)) +
     ylab("Amount ($)") +
     ggtitle("Grants Awared by State") +
     scale_y_continuous(breaks = seq(0,max_,by=100000))+
     theme_hc()
```

## Data Visualization - Advanced Scatter Plots with Facets
```{r echo = FALSE, warning=FALSE, fig.width=10}
# We can add the layer face_grid(~ model_selected) which will facet the chart by the tranformationl mode
qplot(Region, award_amount_2010.2011, data = na.omit(joined), color = Region) +
     theme(axis.text.x = element_text(angle = 70, hjust = 1, vjust = 1)) +
     ylab("Amount ($)") +
     ggtitle("Grants Awared by State") +
     scale_y_continuous(breaks = seq(0,max_,by=100000))+
     theme_hc()  +
     facet_grid(~ model_selected)
```

## Data Visualization - Bar plot
```{r echo = FALSE, warning=FALSE, fig.width=10}
# Now let's make a bar plot of the top ten states with the most funding.
bar_data = head(ordered_states,15)
ggplot(bar_data, aes(state, total, fill = state)) +
      geom_bar(stat = "identity") +
      theme_hc()
```

## Data Visualization - Histogram
```{r echo = FALSE, warning=FALSE, fig.width=10}
qplot(joined$award_amount_2010.2011, geom="histogram") +
     xlab("Grant Amount ($)") +
     theme_hc()
```

## Data Visualization - Histogram with Facet
```{r echo = FALSE, warning=FALSE, fig.width=10}
ggplot(na.omit(joined), aes(x =award_amount_2010.2011)) +
  geom_histogram() + 
  facet_grid( ~model_selected) +
  theme_hc()
```


## Data Visualization - State Bins
```{r echo = FALSE, warning=FALSE, fig.width=10}
statebins_continuous(joined, "state", "award_amount_2010.2011", plot_title = "Heatmap of Grants Awarded by State", )
```

